---
title: "Socio-cultural practices may have affected sex differences in stature in Early Neolithic Europe"
output: html_document
date: '2023-07-20'
---

```{r setup, include=FALSE}
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggvenn)
library(ggplot2)
library(stringr)
library(reshape2)
library(ggsignif)
library(ggbeeswarm)
library(RColorBrewer)
#library(data.table)
library(forcats)
library(BayesFactor)

source("lbkFunctions.R")
## read in the data
compiled <- read.csv("SupplementaryTable1.csv")

## making sure all the data classes are correct after importing them
compiled$nealeClump_prs <- as.numeric(compiled$nealeClump_prs)
compiled$siblingClump_prs <- as.numeric(compiled$siblingClump_prs)
compiled$nealeLDpred_prs <- as.numeric(compiled$nealeLDpred_prs)
compiled$siblingLDpred_prs <- as.numeric(compiled$siblingLDpred_prs)
compiled$method <- as.factor(compiled$method)
compiled$Sex <- as.factor(compiled$Sex)
compiled$Country <- as.factor((compiled$Country))
compiled$Region <- as.factor(compiled$Region)
compiled$Region <- factor(compiled$Region, levels=c("North Central", "South Central", "balkan", "mediterranean", "HG", "anatoliaEF"))
compiled$Region2 <- ifelse(compiled$Region == "North Central" | compiled$Region == "South Central", "Central", as.character(compiled$Region))
compiled$Region2 <- factor(compiled$Region2, levels=c("Central", "balkan", "mediterranean", "anatoliaEF", "HG"))
compiled$CO <- as.factor(compiled$CO)
compiled$LEH <- as.factor(compiled$LEH)
compiled$PH <- as.factor(compiled$PH)
compiled$whgProp <- as.numeric(compiled$whgProp)

## create a dataset without the Mesolithic Hunter-Gatherers and Anatolian Farmers
neoOnly <- compiled[compiled$Region != "HG" & compiled$Region != "anatoliaEF",]

##color palettes for plotting
palette2 <- c("#1B9E77", "#82d682", "#D95F02", "#1275A3", "#BB83A1", "#A8DFF1") ## including HG and EF colors
palette3 <- c("#1B9E77", "#82d682", "#D95F02", "#1275A3", "grey75") ## no HG or EF colors, modern samples in grey
palette4 <- c("#1B9E77", "#82d682", "#D95F02", "#1275A3", "#BB83A1", "#A8DFF1", "grey75") ## including HG and EF colors, modern samples in grey
paleoColors <- c("#495B76", "#609FA6") ## paleopathology presence/absence

```

```{r statTests, echo=FALSE, warning=FALSE}
## dataframes with single sex data
female <- subset(neoOnly, Sex=="F")
male <- subset(neoOnly, Sex=="M")

lm1 <- lm(femur_estimated ~ Region, data=male)
## change the reference level for females to compare regions with the Mediterranean
female$Region <- factor(female$Region, levels=c("mediterranean", "North Central", "South Central", "balkan"))
lm2 <- lm(femur_estimated ~ Region, data=female)
lm3 <- lm(femur_estimated ~ Sex + Region, data=neoOnly)

ttest1 <- pairwise.t.test(neoOnly$nealeClump_prs, neoOnly$Region)
ttest4 <- pairwise.t.test(neoOnly$nealeLDpred_prs, neoOnly$Region)
ttest2 <- pairwise.t.test(neoOnly$siblingLDpred_prs, neoOnly$Region)
ttest3 <- pairwise.t.test(neoOnly$siblingClump_prs, neoOnly$Region)
lm5 <- lm(nealeClump_prs ~ Sex * Region, data=neoOnly)

lm6 <- lm(d13C ~ Sex * Region, data=neoOnly)
lm7 <- lm(d15N ~ Sex * Region, data=neoOnly)
## we report results of the Type II ANOVA in the paper
aov1 <- aov(lm(d15N ~ Sex * Region, data=neoOnly))
aov2 <- aov(lm(d15N ~ Region * Sex, data=neoOnly))

lm8 <- lm(femur_estimated ~ Region, data=female)
lm9 <- lm(femur_estimated ~ Region, data=male)
lm10 <- lm(femur_estimated ~ Sex + nealeClump_prs + Region, data = neoOnly)

lm11 <- lm(femur_estimated ~ Sex + Region + d15N, data = neoOnly)
lm12 <- lm(femur_estimated ~ Sex + Region + d13C, data = neoOnly)

lm13 <- lm(femur_estimated ~ Sex * LEH + Region, data=neoOnly)
glm1 <- glm(LEH ~ femur_estimated + Sex + Region, data=neoOnly, family="binomial")
lm14 <- lm(femur_estimated ~ LEH + Region, data=female)
lm15 <- lm(femur_estimated ~ LEH + Region, data=male)
glm2 <- glm(CO ~ femur_estimated + Sex + Region, data=neoOnly, family="binomial")
lm16 <- lm(femur_estimated ~ Sex * CO + Region, data=neoOnly)
lm17 <- lm(femur_estimated ~ Sex + PH + Region, data=neoOnly)
glm3 <- glm(PH ~ femur_estimated + Sex + Region, data=neoOnly, family="binomial")

lm18 <- lm(whgProp ~ Region, data=compiled)
#function for calculating standard error
se <- function(x) sd(x, na.rm=TRUE)/sqrt(length(x))
mean_whgprop_north <- mean(compiled[compiled$Region == "North Central", "whgProp"], na.rm=TRUE)
se_whgprop_north <- se(compiled[compiled$Region == "North Central", "whgProp"])
mean_whgprop_south <- mean(compiled[compiled$Region == "South Central", "whgProp"], na.rm=TRUE)
se_whgprop_south <- se(compiled[compiled$Region == "South Central", "whgProp"])
mean_whgprop_balkan <- mean(compiled[compiled$Region == "balkan", "whgProp"], na.rm=TRUE)
se_whgprop_balkan <- se(compiled[compiled$Region == "balkan", "whgProp"])
mean_whgprop_med <- mean(compiled[compiled$Region == "mediterranean", "whgProp"], na.rm=TRUE)
se_whgprop_med <- se(compiled[compiled$Region == "mediterranean", "whgProp"])
lm19 <- lm(nealeClump_prs ~ Region, data=compiled)
lm20 <- lm(femur_estimated ~ Sex + whgProp, data=compiled)
lm21<- lm(siblingLDpred_prs ~ Region, data=compiled)

```

``` {r sexDim, echo=FALSE, warning=FALSE}
## estimate statures using the estimated femora and Ruff et al (2012) stature method
statures <- statCalc(compiled, "femur_estimated", "Sex")

## table of sexual dimorphism ratios in femur length (femDiff)
femDiff <- aggregate(femur_estimated ~ Sex + Region, data=statures, mean)
femDiffF <- subset(femDiff, Sex=="F")
femDiffM <- subset(femDiff, Sex=="M")
femDiff <- femDiffF[,2:3]
femDiff$MtoF_femur <- NA
femDiff$percentDiff <- NA
for(i in 1:nrow(femDiffF)){
   femDiff[i,3] <- femDiffM[i,3]/femDiffF[i,3]
   femDiff[i,4] <- round((femDiffM[i,3] - femDiffF[i,3])/femDiffF[i,3] * 100, 2)
}
femDiff <- femDiff[,-2]

## table of sexual dimorphism ratios in stature (femDiff)
sexDim <- aggregate(stature_est~ Sex + Region, data=statures, mean)
sexDimF <- subset(sexDim, Sex=="F")
sexDimM <- subset(sexDim, Sex=="M")
sexDim <- sexDimF[,2:3]
sexDim$MtoF_femur <- NA
sexDim$percentDiff <- NA
for(i in 1:nrow(femDiffF)){
   sexDim[i,3] <- sexDimM[i,3]/sexDimF[i,3]
   sexDim[i,4] <- round((sexDimM[i,3] - sexDimF[i,3])/sexDimF[i,3] * 100, 2)
}

```

```{r bayesFactors, echo=FALSE, warning=FALSE}

# Nitrogen is generally elevated in males compared to females, but only significant in the Med lm7
# model for results in the paper is d15N ~ Sex * Region
# lm7_na.omit <- na.omit(neoOnly[,c("d15N", "Sex", "Region")])
# bf7.1a <- lmBF(d15N ~ Sex + Region, data=lm7_na.omit)
# bf7.1b <- lmBF(d15N ~ Sex, data=lm7_na.omit)
# bf7.1 <- bf7.1a/bf7.1b

# Males in both regions have qualitatively higher nitrogen, but interaction effect is not significant
lm7_na.omit <- na.omit(neoOnly[,c("d15N", "Sex", "Region")])
bf7a <- lmBF(d15N ~ Sex * Region, data=lm7_na.omit)
bf7b <- lmBF(d15N ~ Sex + Region, data=lm7_na.omit)
bf7 <- bf7a/bf7b

# *Effect of PRS on femur length is just significant* lm10
lm10_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "nealeClump_prs", "Region")])
bf10a <- lmBF(femur_estimated ~ Sex + nealeClump_prs + Region, data = lm10_na.omit)
bf10b <- lmBF(femur_estimated ~ Sex + Region, data = lm10_na.omit)
bf10 <- bf10a/bf10b

# 13C are the same between sexe lm6
lm6_na.omit <- na.omit(neoOnly[,c("d13C", "Sex", "Region")])
bf6a <- lmBF(d13C ~ Sex * Region, data = lm6_na.omit)
bf6b <- lmBF(d13C ~ Sex + Region, data = lm6_na.omit)
bf6 <- bf6a/bf6b

# No statistical relationship between femur length and 15N lm11
lm11_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "d15N", "Region")])
bf11a <- lmBF(femur_estimated ~ Sex + Region + d15N, data = lm11_na.omit)
bf11b <- lmBF(femur_estimated ~ Sex + Region, data = lm11_na.omit)
bf11 <- bf11a/bf11b

# No statistical relationship between femur length and 13C lm12
lm12_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "d13C", "Region")])
bf12a <- lmBF(femur_estimated ~ Sex + Region + d13C, data = lm12_na.omit)
bf12b <- lmBF(femur_estimated ~ Sex + Region, data = lm12_na.omit)
bf12 <- bf12a/bf12b

# No difference between incidence of LEH in males vs females glm1
glm1_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "LEH", "Region")])
glm1_na.omit$LEH <- as.numeric(glm1_na.omit$LEH)
bfg1a <- lmBF(LEH ~ femur_estimated + Sex + Region, data = glm1_na.omit)
bfg1b <- lmBF(LEH ~ femur_estimated + Region, data = glm1_na.omit)
bfg1 <- bfg1a/bfg1b

# Interaction effect between sex and LEH on femur length is not significant lm13
lm13_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "LEH", "Region")])
bf13a <- lmBF(femur_estimated ~ Sex * LEH + Region, data = lm13_na.omit)
bf13b <- lmBF(femur_estimated ~ Sex + LEH + Region, data = lm13_na.omit)
bf13 <- bf13a/bf13b

# Females with LEH have significantly shorter femora than females without lm14
lm14_na.omit <- na.omit(female[,c("femur_estimated", "LEH", "Region")])
bf14a <- lmBF(femur_estimated ~ LEH + Region, data = lm14_na.omit)
bf14b <- lmBF(femur_estimated ~ Region, data = lm14_na.omit)
bf14 <- bf14a/bf14b

# Males with LEH have the same femur length as those without lm15
lm15_na.omit <- na.omit(male[,c("femur_estimated", "LEH", "Region")])
bf15a <- lmBF(femur_estimated ~ LEH + Region, data = lm15_na.omit)
bf15b <- lmBF(femur_estimated ~ Region, data = lm15_na.omit)
bf15 <- bf15a/bf15b

# Cribra has no statistical relationship with femur length lm16
lm16_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "CO", "Region")])
bf16a <- lmBF(femur_estimated ~ Sex * CO + Region, data = lm16_na.omit)
bf16b <- lmBF(femur_estimated ~ Sex + CO + Region, data = lm16_na.omit)
bf16 <- bf16a/bf16b

# Frequencies of porotic hyperostosis are not different between North and South glm3
glm3_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "PH", "Region")])
glm3_na.omit$PH <- as.numeric(glm3_na.omit$PH)
bfg3a <- lmBF(PH ~ femur_estimated + Sex + Region, data = glm3_na.omit)
bfg3b <- lmBF(PH ~ femur_estimated + Sex, data = glm3_na.omit)
bfg3 <- bfg3a/bfg3b

# Porotic hyperostosis is not related to femur length lm17
lm17_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "PH", "Region")])
bf17a <- lmBF(femur_estimated ~ Sex + PH + Region, data = lm17_na.omit)
bf17b <- lmBF(femur_estimated ~ Sex + Region, data = lm17_na.omit)
bf17 <- bf17a/bf17b

#*Proportions of WHG are weakly associated with femur length* lm20
lm20_na.omit <- na.omit(neoOnly[,c("femur_estimated", "Sex", "whgProp")])
bf20a <- lmBF(femur_estimated ~ Sex + whgProp, data = lm20_na.omit)
bf20b <- lmBF(femur_estimated ~ Sex, data = lm20_na.omit)
bf20 <- bf20a/bf20b

```

Samantha L Cox, Nicole Nicklisch, Michael Francken, Joachim Wahl, Harald Meller, Wolfgang Haak, Kurt W Alt, Eva Rosenstock, Iain Mathieson

## Results
### Distribution of stature, polygenic scores, and stable isotope values

- Med males are significantly shorter than Central or Balkans (p=`r round(summary(lm1)$coefficients[4,4], 4)`, beta=`r round(summary(lm1)$coefficients[4,1], 3)`, se=`r round(summary(lm1)$coefficients[4,2], 4)`)
- Differences between Central and Balkan male femora are very small (North vs. South Central: p=`r round(summary(lm1)$coefficients[2,4], 4)`, beta=`r round(summary(lm1)$coefficients[2,1], 3)`, se=`r round(summary(lm1)$coefficients[2,2], 4)`; North Central vs Balkan: p=`r round(summary(lm1)$coefficients[3,4], 4)`, beta=`r round(summary(lm1)$coefficients[3,1], 3)`, se=`r round(summary(lm1)$coefficients[3,2], 4)`)  
- Female femora in the Med are similar to those in South Central and Balkan populations (Med vs. South Central: p=`r round(summary(lm2)$coefficients[3,4], 4)`, beta=`r round(summary(lm2)$coefficients[3,1], 3)`, se=`r round(summary(lm2)$coefficients[3,2], 4)`; Med vs Balkan: p=`r round(summary(lm2)$coefficients[4,4], 4)`, beta=`r round(summary(lm2)$coefficients[4,1], 3)`, se=`r round(summary(lm2)$coefficients[4,2], 4)`)
- Female femora are substantially shorter in the North Central population (p=`r round(summary(lm2)$coefficients[2,4], 4)`, beta=`r round(summary(lm2)$coefficients[2,1], 3)`, se=`r round(summary(lm2)$coefficients[2,2], 4)`)
- Differences between male and female femur lengths are highly significant in all populations (p=`r round(summary(lm3)$coefficients[2,4], 4)`, beta=`r round(summary(lm3)$coefficients[2,1], 3)`, se=`r round(summary(lm3)$coefficients[2,2], 4)`)

- PRS for height are very similar between all populations (min. p=`r round(ttest1$p.value[1,1], 4)`, DF=`r nrow(neoOnly[!(is.na(neoOnly$nealeClump_prs)),]) - 2`)
- PRS constructed with LDpred show Med individuals are shorter (max. p=`r round(ttest4$p.value[3,1], 4)`, DF=`r nrow(neoOnly[!(is.na(neoOnly$nealeClump_prs)),]) - 2`)
- PRS constructed with sibling summary stats show similar PRS in all populations using both LDpred (min. p-value between Balkans and Mediterranean =`r round(ttest2$p.value[3,3], 4)`, DF=`r nrow(neoOnly[!(is.na(neoOnly$siblingLDpred_prs)),]) - 2`) and clumping/thresholding construction methods (min. p-value between Balkans and Mediterranean =`r round(ttest3$p.value[3,3], 4)`, DF=`r nrow(neoOnly[!(is.na(neoOnly$siblingClump_prs)),]) - 2`)
- No significant differences between male and female PRS in any population (p=`r round(summary(lm5)$coefficients[4,4], 4)`, beta=`r round(summary(lm5)$coefficients[4,1], 3)`, se=`r round(summary(lm5)$coefficients[4,2], 4)`)

- Med and Balkan groups are significantly distinct from Central Europe in 13C (vs. Balkans p = `r round(summary(lm6)$coefficients[4,4], 4)`, beta=`r round(summary(lm6)$coefficients[4,1], 3)`, se=`r round(summary(lm6)$coefficients[4,2], 4)`; vs. Med p = `r round(summary(lm6)$coefficients[5,4], 4)`, beta=`r round(summary(lm6)$coefficients[5,1], 3)`, se=`r round(summary(lm6)$coefficients[5,2], 4)`)
- North Central is different from South Central and Balkan groups 15N (vs. South Central p = `r round(summary(lm7)$coefficients[3,4], 4)`, beta=`r round(summary(lm7)$coefficients[3,1], 3)`, se=`r round(summary(lm7)$coefficients[3,2], 4)`; vs. Balkans p = `r round(summary(lm7)$coefficients[4,4], 4)`, beta=`r round(summary(lm7)$coefficients[4,1], 3)`, se=`r round(summary(lm7)$coefficients[4,2], 4)`)
- Nitrogen is generally elevated in males compared to females (F=`r round(summary(aov2)[[1]][["F value"]][2], 1)`, DF=`r summary(aov2)[[1]][["Df"]][2]`,`r summary(aov2)[[1]][["Df"]][4]`, p=`r round(summary(aov2)[[1]][["Pr(>F)"]][2], 3)`), and N values differ across regions (F=`r round(summary(aov1)[[1]][["F value"]][2], 1)`, DF=`r summary(aov1)[[1]][["Df"]][2]`,`r summary(aov1)[[1]][["Df"]][4]`, p=`r round(summary(aov1)[[1]][["Pr(>F)"]][2], 3)`) with a significant interaction term (F=`r round(summary(aov1)[[1]][["F value"]][3], 1)`, DF=`r summary(aov1)[[1]][["Df"]][3]`,`r summary(aov1)[[1]][["Df"]][4]`, p=`r round(summary(aov1)[[1]][["Pr(>F)"]][3], 3)`)

### Patterns of non-genetic factors in Central Europe
- Female femora in the North are significantly shorter than in the South (p=`r round(summary(lm8)$coefficients[2,4], 4)`; beta=`r summary(lm8)$coefficients[2,1]`, se=`r round(summary(lm8)$coefficients[2,2], 4)`) 
- Male femora in the North are very similar to those in the South (p=`r round(summary(lm9)$coefficients[2,4], 4)`; beta=`r summary(lm9)$coefficients[2,1]`, se=`r round(summary(lm9)$coefficients[2,2], 4)`)
- sexual dimorphism ratios (femora)
``` {r femDimorphism, echo=FALSE, warning=FALSE}
print(femDiff)
```

- sexual dimorphism ratios (stature)
``` {r statDimorphism, echo=FALSE, warning=FALSE}
print(sexDim)
```

- Effect of PRS on femur length is just significant (p=`r round(summary(lm10)$coefficients[3,4], 4)`; beta=`r summary(lm10)$coefficients[3,1]`, se=`r round(summary(lm10)$coefficients[3,2], 4)`, bf=`r bf10@bayesFactor$bf`) 
- South Central Europe has higher 15N than North (p=`r round(summary(lm7)$coefficients[3,4], 4)`; beta=`r summary(lm7)$coefficients[3,1]`, se=`r round(summary(lm7)$coefficients[3,2], 4)`)
- South Central Europe has lower 13C than North (p=`r round(summary(lm6)$coefficients[3,4], 4)`; beta=`r summary(lm6)$coefficients[3,1]`, se=`r round(summary(lm6)$coefficients[3,2], 4)`)
- Males in both regions have qualitatively higher nitrogen, but interaction effect is not significant (p=`r round(summary(lm7)$coefficients[6,4], 4)`; beta=`r summary(lm7)$coefficients[6,1]`, se=`r round(summary(lm7)$coefficients[6,2], 4)`, bf=`r bf7@bayesFactor$bf`) 
- 13C are the same between sexes (p=`r round(summary(lm6)$coefficients[6,4], 4)`; beta=`r summary(lm6)$coefficients[6,1]`, se=`r round(summary(lm6)$coefficients[6,2], 4)`, bf=`r bf6@bayesFactor$bf`)
- No statistical relationship between femur length and 15N (p=`r round(summary(lm11)$coefficients[4,4], 4)`; beta=`r summary(lm11)$coefficients[4,1]`, se=`r round(summary(lm11)$coefficients[4,2], 4)`, bf=`r bf11@bayesFactor$bf`) 
- No statistical relationship between femur length and 13C (p=`r round(summary(lm12)$coefficients[4,4], 4)`; beta=`r summary(lm12)$coefficients[4,1]`, se=`r round(summary(lm12)$coefficients[4,2], 4)`, bf=`r bf12@bayesFactor$bf`)

- Femur length is significantly associated with presence of LEH (p=`r round(summary(lm13)$coefficients[3,4], 4)`; beta=`r summary(lm13)$coefficients[3,1]`, se=`r round(summary(lm13)$coefficients[3,2], 4)`)
- Northern individuals more likely to have LEH than Southern (p=`r round(summary(glm1)$coefficients[4,4], 4)`; beta=`r summary(glm1)$coefficients[4,1]`, se=`r round(summary(glm1)$coefficients[4,2], 4)`)
- No difference between incidence of LEH in males vs females (p=`r round(summary(glm1)$coefficients[3,4], 4)`; beta=`r summary(glm1)$coefficients[3,1]`, se=`r round(summary(glm1)$coefficients[3,2], 4)`, bf=`r bfg1@bayesFactor$bf`) 
- Interaction effect between sex and LEH on femur length is not significant (p=`r round(summary(lm13)$coefficients[5,4], 4)`; beta=`r summary(lm13)$coefficients[5,1]`, se=`r round(summary(lm13)$coefficients[5,2], 4)`, bf=`r bf13@bayesFactor$bf`) 
- Females with LEH have significantly shorter femora than females without (p=`r round(summary(lm14)$coefficients[2,4], 4)`; beta=`r summary(lm14)$coefficients[2,1]`, se=`r round(summary(lm14)$coefficients[2,2], 4)`, bf=`r bf14@bayesFactor$bf`)
- Males with LEH have the same femur length as those without (p=`r round(summary(lm15)$coefficients[2,4], 4)`; beta=`r summary(lm15)$coefficients[2,1]`, se=`r round(summary(lm15)$coefficients[2,2], 4)`, bf=`r bf15@bayesFactor$bf`)
- Incidence of cribra orbitalia is significantly higher in the North compared to South (p=`r round(summary(glm2)$coefficients[4,4], 4)`; beta=`r summary(glm2)$coefficients[4,1]`, se=`r round(summary(glm2)$coefficients[4,2], 4)`)
- Cribra has no statistical relationship with femur length (p=`r round(summary(lm16)$coefficients[3,4], 4)`; beta=`r summary(lm16)$coefficients[3,1]`, se=`r round(summary(lm16)$coefficients[3,2], 4)`, bf=`r bf16@bayesFactor$bf`)
- Frequencies of porotic hyperostosis are not different between North and South (p=`r round(summary(glm3)$coefficients[4,4], 4)`; beta=`r summary(glm3)$coefficients[4,1]`, se=`r round(summary(glm3)$coefficients[4,2], 4)`, bf=`r bfg3@bayesFactor$bf`)
- Porotic hyperostosis is not related to femur length (p=`r round(summary(lm17)$coefficients[3,4], 4)`; beta=`r summary(lm17)$coefficients[3,1]`, se=`r round(summary(lm17)$coefficients[3,2], 4)`, bf=`r bf17@bayesFactor$bf`) 

### Patterns of genetic ancestry in the Mediterranean
- Proportions of WHG ancestry are significantly higher in the Med as compared to Central Europe (p=`r round(summary(lm18)$coefficients[4,4], 4)`; beta=`r summary(lm17)$coefficients[4,1]`, se=`r round(summary(lm17)$coefficients[4,2], 4)`)
- Average proportion of WHG ancestry in Med (mean=`r round(mean_whgprop_med*100, 2)`%, se=`r round(se_whgprop_med, 4)*100`%), Balkans (mean=`r round(mean_whgprop_balkan*100, 2)`%, se=`r round(se_whgprop_balkan, 4)*100`%), South Central (mean=`r round(mean_whgprop_south*100, 2)`%, se=`r round(se_whgprop_south, 4)*100`%), North Central (mean=`r round(mean_whgprop_north*100, 2)`%, se=`r round(se_whgprop_north, 4)*100`%)
- Clumping/thresholding PRS is lowest in WHG (p=`r round(summary(lm19)$coefficients[5,4], 4)`; beta=`r summary(lm19)$coefficients[5,1]`, se=`r round(summary(lm19)$coefficients[5,2], 4)`)
- Anatolian clumping/thresholding PRS is similar to Balkans and Central Europeans (p=`r round(summary(lm19)$coefficients[6,4], 4)`; beta=`r summary(lm19)$coefficients[6,1]`, se=`r round(summary(lm19)$coefficients[6,2], 4)`)
- Proportions of WHG are weakly associated with femur length (p=`r round(summary(lm20)$coefficients[3,4], 4)`; beta=`r summary(lm20)$coefficients[3,1]`, se=`r round(summary(lm20)$coefficients[3,2], 4)`, bf=`r bf20@bayesFactor$bf`)
- WHG PRS is significantly greater than other groups when computed with LDpred and sibling summary stats (p=`r round(summary(lm21)$coefficients[5,4], 4)`; beta=`r summary(lm21)$coefficients[5,1]`, se=`r round(summary(lm21)$coefficients[5,2], 4)`)

## Fig 2
```{r Fig2, echo=FALSE, out.width="50%", warning=FALSE}

## create the sample maps
europe <- ne_countries(scale = "medium", returnclass = "sf", continent="Europe")
map <- ggplot(data = europe) +
  geom_sf() +
  xlab("Longitude") + ylab("Latitude") +
  geom_point(data=neoOnly[neoOnly$dna_ID != "",], aes(x=Longitude, y=Latitude, col=Region)) +
  scale_color_manual(values=palette2) +
   coord_sf(xlim = c(-10, 30), ylim = c(35, 55))+
   theme_light() +
  ggtitle("2A) Map of genetic sites")
print(map)

map2 <- ggplot(data = europe) +
  geom_sf() +
  xlab("Longitude") + ylab("Latitude") +
  geom_point(data=neoOnly[neoOnly$metric_ID != "",], aes(x=Longitude, y=Latitude, col=Region)) +
  scale_size_continuous(breaks=c(10, 50, 100)) +
   scale_color_manual(values=palette2) +
  coord_sf(xlim = c(-10, 30), ylim = c(35, 55))+
   theme_light() +
  ggtitle("2B) Map of skeletal sites")
print(map2)

## import the smartPCA data
impPCA <- read.csv("impPCA_neoData.csv")
impPCA$Region <- as.factor(impPCA$Region)
## rearrange factor levels
impPCA$Region <- factor(impPCA$Region, levels=c("North Central", "South Central", "balkan", 
                                                          "mediterranean", "modern"))
## make the imputed PCA plot
ggplot(data=impPCA, aes(x=PC1, y=-PC2, col=Region, stroke=0)) +
    geom_point() +
    scale_color_manual(values=palette3) +
    ggtitle("2C) Genetic Ancestry: projected PCA (imputed)") +
    theme_light()

## plot of all samples by date in years BP
neoOnly$type <- ifelse(neoOnly$dna_ID != "", "genetic", "skeletal")
neoOnly$type <- ifelse(neoOnly$type == "skeletal" & neoOnly$isotope_ID != "", "isotopes", "skeletal")
ggplot(data=neoOnly, aes(x=Region, y=Date, col=Region)) + 
  geom_beeswarm(dodge.width=0.95) +
  scale_color_manual(values=palette2) +
  facet_wrap(~type) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("2D) Samples by Date (yearsBP)")
```

## Fig 3
```{r fig3, echo=FALSE, out.width="50%", warning=FALSE}

ggplot(data=neoOnly, aes(y=femur_estimated, x=Region, col=Region, shape=Sex, fill=Sex, stroke=0)) +
   geom_beeswarm(dodge.width=0.75) +
   stat_summary(fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5, position="dodge") +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   theme(axis.text.x = element_text(angle = 90)) +
   ggtitle("3A) Femur length of Neolithic Europeans")

ggplot(data=neoOnly, aes(y=nealeClump_prs, x=Region, col=Region, shape=Sex, stroke=0)) +
   geom_beeswarm(dodge.width = 0.75) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5, position="dodge") +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   geom_signif(comparisons=list(c("Northern LBK", "mediterranean")), test=t.test) + 
   ggtitle("3B) PRS of Neolithic Europeans")

ggplot(data=neoOnly, aes_string(y="d15N", x="d13C", col="Region", stroke=0)) +
   geom_point() +
   theme(axis.text.x = element_text(angle = 90)) +
   guides(scale="none") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("3C: d13C and d15N")

ggplot(data=neoOnly[!(is.na(neoOnly$Sex)),], aes(x=Region, y=d15N, shape=Sex, col=Region, stroke=0)) +
   #geom_boxplot() +
   # guides(size=FALSE) +
   geom_beeswarm(dodge.width=0.75) +
   stat_summary(fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5, position="dodge") +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("3D: sex differences in d15N")

```

## Fig 4
```{r fig4, echo=FALSE, out.width="50%", warning=FALSE}
## subset data for plotting just Central European samples
isoplot <- subset(compiled, Region == "North Central" | Region == "South Central")
## remove individuals which do not have stable isotopic data
isoplot <- subset(isoplot, !(is.na(isoplot$d15N)) | !(is.na(isoplot$d13C)))
## remove factor levels from the full dataset which are not present in this subset
isoplot$Region <- droplevels(isoplot$Region)
isoplot$Sex <- droplevels(isoplot$Sex)
## reshape the data for easier ggplotting
isoplot <- reshape2::melt(isoplot, id.vars=c("isotope_ID", "Sex", "Region"), measure.vars=c("d15N", "d13C"))

ggplot(data=isoplot[!is.na(isoplot$Sex),], aes(y=value, x=Region, col=Sex, fill=Sex, stroke=0)) +
   # geom_boxplot() +
   geom_beeswarm(dodge.width=0.75) +
   scale_color_manual(values=palette2) +
   stat_summary(fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5, position="dodge") +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   geom_signif(comparisons=list(c("LBK", "mediterranean")), test=t.test) + 
   facet_wrap(~variable, scales="free_y") +
   theme_light() +
   ggtitle("4A) LBK isotopes by sex")


LEHplot <- compiled[!is.na(compiled$LEH),]
LEHplot <- subset(LEHplot, Sex == "M" | Sex == "F")
LEHplot$Region <- droplevels(LEHplot$Region)
LEHplot$Sex <- droplevels(LEHplot$Sex)

ggplot(data=LEHplot, aes(x=Region, fill=LEH, stroke=0)) +
   geom_bar(position="fill") +
   facet_grid(~Sex) +
   scale_y_continuous(labels = scales::percent) +
   scale_fill_manual(values=paleoColors) +
   guides(size=FALSE) +
   theme_light() +
   ggtitle("4B) LBK femur and linear enamel hypoplasia")

ggplot(data=LEHplot[!is.na(LEHplot$Sex) & LEHplot$LEH != "",], aes(y=femur_estimated, x=Sex, fill=LEH, shape=Sex, stroke=0)) +
   geom_beeswarm(dodge.width=0.75, aes(col=LEH)) +
  facet_wrap(~Region) +
   stat_summary(fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5, position="dodge") +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=paleoColors) +
   theme_light() +
   ggtitle("4C) LBK femur and linear enamel hypoplasia")

```

## Fig 5
```{r fig5, echo=FALSE, out.width="50%", warning=FALSE}
## read in data from smartPCA including WHG and Anatolian Farmers
impPCAall <- read.csv("impPCA_allData.csv")
impPCAall$Region <- as.factor(impPCAall$Region)
## reorder the Region factor levels
impPCAall$Region <- factor(impPCAall$Region, levels=c("North Central", "South Central", "balkan", "mediterranean", "HG", "anatoliaEF", "modern"))

ggplot(data=impPCAall, aes(x=PC1, y=-PC2, col=Region, stroke=0)) +
    geom_point() +
    scale_color_manual(values=palette4) +
    ggtitle("5A) Genetic Ancestry: projected PCA (imputed)") +
    theme_light()



#####
## read in the ADMIXTURE results 
mix <- read.table("admixtureData.txt", header=TRUE)
## reorder factor levels for plotting
mix$Region <- factor(mix$Region, levels=c("North Central", "South Central", "balkan", 
                                                "mediterranean", "HG", "anatoliaEF"))
## order dataframe by admixture proportion for plotting
mix$ID <-fct_reorder(mix$ID, mix$EF)
## reshape admixture data for plotting
toplotMix <- reshape2::melt(mix, c("ID", "Region", "Sex"))
toplotMix$variable <- factor(toplotMix$variable, levels=c("EF", "WHG"))

ggplot(data=toplotMix, aes(x=ID, y=value, fill=variable, col=variable), position="stack") +
  geom_col() +
  facet_grid(~Region, scales="free_x", space="free_x") + 
  scale_fill_manual(values=palette4[5:6]) +
  scale_color_manual(values=palette4[5:6]) +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  ggtitle("5B) ADMIXTURE")

ggplot(data=compiled, aes(y=nealeClump_prs, x=Region, col=Region, stroke=0)) +
   geom_beeswarm(cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("5C) PRS of Neolithic Europeans, Hunter-Gatherers, and Anatolia Farmers")

ggplot(data=compiled, aes(y=siblingClump_prs, x=Region, col=Region, stroke=0)) +
   geom_beeswarm(cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("5D) Sibling-based PRS of Neolithic Europeans, Hunter-Gatherers, and Anatolia Farmers")

```

# Supplemenary Figures
## Supp Fig 1
```{r suppFig1, echo=FALSE, out.width="50%", warning=FALSE}
venn <- list("DNA"=which(compiled$dna_ID != ""), "isotopes"=which(compiled$isotope_ID != ""), "metrics"=which(compiled$metric_ID != ""), "pathology"=which(compiled$paleopath_ID != ""))
ggvenn(venn, fill_color = c("#0073C2FF", "#EFC000FF", "#CD534CFF", "grey"), stroke_size = 0.5, show_percentage = FALSE, text_size = 5) +
  scale_x_continuous(expand = expansion(mult = .2)) +
  ggtitle("Supp.Fig. 1: Data set counts")

```
## Supp Fig 2
```{r suppFig2, echo=FALSE, out.width="50%", warning=FALSE}
sexspecific <- read.csv("sex-specific_prs_data.csv")
sexspecific$Region <- factor(sexspecific$Region, levels=c("North Central", "South Central", "balkan", "mediterranean", "HG", "anatoliaEF"))

ggplot(data=sexspecific, aes(y=combo_prs, x=Region, col=Region, shape=Sex, stroke=0)) +
   #geom_boxplot() +
   geom_beeswarm(dodge.width=0.75, cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("A) Neale Lab Combined Sex sumstats")

ggplot(data=sexspecific, aes(y=asMales_prs, x=Region, col=Region, shape=Sex, stroke=0)) +
   geom_beeswarm(dodge.width=0.75, cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("B) Neale Lab Male sumstats")

ggplot(data=sexspecific, aes(y=asFemales_prs, x=Region, col=Region, shape=Sex, stroke=0)) +
   geom_beeswarm(dodge.width=0.75, cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("C) Neale Lab Female sumstats")

ggplot(data=sexspecific, aes(y=sexspec_prs, x=Region, col=Region, shape=Sex, stroke=0)) +
   geom_beeswarm(dodge.width=0.75, cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("D) Neale Lab sex-specific sumstats")


```

## Supp Fig 3
```{r suppFig3, echo=FALSE, warning=FALSE}
ggplot(data=compiled, aes(y=siblingLDpred_prs, x=Region, col=Region, stroke=0)) +
   geom_beeswarm(cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   geom_signif(comparisons=list(c("Northern LBK", "mediterranean")), test=t.test) + 
   ggtitle("A) Sibling LDPred2 PRS")

ggplot(data=compiled, aes(y=nealeClump_prs, x=femur_estimated)) +
   geom_point(aes(col=Region)) +
   geom_smooth(method="lm") +
   scale_color_manual(values=palette2) +
   theme_light() +
   ggtitle("B) Sibling PRS vs Femur length")

```

## Supp Fig 4
```{r suppFig4, echo=FALSE, warning=FALSE}
## import data for the unimputed Neolithic PCA
unimpPCA <- read.csv("unimpPCA_neoData.csv")
unimpPCA$Region <- as.factor(unimpPCA$Region)
## reorder the factor levels for plotting
unimpPCA$Region <- factor(unimpPCA$Region, levels=c("North Central", "South Central", "balkan",
                                                          "mediterranean", "modern"))

ggplot(data=unimpPCA, aes(x=-PC1, y=PC2, col=Region, stroke=0)) +
    geom_point() +
    scale_color_manual(values=palette3) +
    labs(x="PC1", y="PC2") +
    ggtitle("A) Genetic Ancestry: projected PCA (unimputed)") +
    theme_light()


## import all unimputed PCA data
unimpPCAall <- read.csv("unimpPCA_allData.csv")
unimpPCAall$Region <- as.factor(unimpPCAall$Region)
## reorder factor levels for plotting
unimpPCAall$Region <- factor(unimpPCAall$Region, levels=c("North Central", "South Central", "balkan", "mediterranean", "HG", "anatoliaEF", "modern"))

ggplot(data=unimpPCAall, aes(x=-PC1, y=PC2, col=Region, stroke=0)) +
    geom_point() +
    scale_color_manual(values=palette4) +
    labs(x="PC1", y="PC2") +
    ggtitle("B) Genetic Ancestry: projected PCA (unimputed)") +
    theme_light()
```

## Supp Fig 5
```{r suppFig5, echo=FALSE, warning=FALSE}

ggplot(data=neoOnly, aes(x=Region, y=nealeLDpred_prs, col=Region, shape=Sex, stroke=0)) +
   geom_beeswarm(cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
    scale_color_manual(values=palette3) +
    ggtitle("PRS using Neale Lab summary stats and LDpred2") +
    theme_light()
```

## Supp Fig 6
```{r suppFig6, echo=FALSE, warning=FALSE}
## import data for the LBK herbivore isotopes
animals <- read.csv("Animal_isotopes.csv")
## remove the regional means for plotting
animals <- animals[-c(104:105),]
## reorder the factor levels for plotting
animals$region <- as.factor(animals$region)

ggplot(data=animals, aes(x=d13C, y=d15N, col=region, stroke=0)) +
    geom_point() +
    scale_color_manual(values=palette3) +
    labs(x="delta 13C", y="delta 15N") +
    ggtitle("Difference in herbivore baseline isotopes for Central Europe") +
    theme_light()
```

## Supp Fig 7
```{r suppFig7, echo=FALSE, warning=FALSE}
dateRestrict <- neoOnly[neoOnly$Date > 6900 & neoOnly$Date < 7200,]

ggplot(data=dateRestrict, aes(x=Region, y=nealeClump_prs, col=Region, shape=Sex, stroke=0)) +
   geom_beeswarm(dodge.width=0.75, cex=2) +
   stat_summary( fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5) +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
    scale_color_manual(values=palette3) +
    ggtitle("A) PRS restricted to 6900-7200BP") +
    theme_light()

ggplot(data=dateRestrict, aes(y=femur_estimated, x=Region, col=Region, shape=Sex, fill=Sex, stroke=0)) +
   geom_beeswarm(dodge.width=0.75) +
   stat_summary(fun = "mean",
               fun.max = function(x) mean(x) + sd(x) / sqrt(length(x)),
               fun.min = function(x) mean(x) - sd(x) / sqrt(length(x)), 
               geom = "point", col = "black", width=.5, position="dodge") +
   stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", col = "black", position="dodge") +
   scale_color_manual(values=palette2) +
   theme_light() +
   theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("B) Femora restricted to 6900-7200BP")

```
